Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-13 Manuscript/MPESubmission/Revision/response.tex

\item{\textcolor{blue}{\textbf{line 540:} replace 'a lot' with 'larger proportions of missing data', a lot seems too colloquial for this kind of article. }}

We modified the sentence as suggested: "they have a larger proportion of missing data" (line @@@).

\item{\textcolor{blue}{\textbf{lines 542-547:} These seem like the most important practical results and should be more strongly emphasized. These are very important results for practitioners and the authors should try to make it more clear to the reader that this is a key take-home message.}}

We modified the sentence to:

"Surprisingly, the number of missing living taxa with morphological data ($M_{L}$) and the overall number of missing morphological characters ($M_{C}$), have a bigger effect than the amount of missing data for the fossil taxa ($M_{F}$).
It seems therefore more important to integrate a maximum of living taxa with morphological data as well as a maximum number of characters.
Additionally, when both $M_{L}$ and $M_{C}$ reach 50\% of missing data, the matrix does not contain enough phylogenetic information for the fossil taxa to be placed with confidence in the tree (Figure 6)."

%Original: The number of missing living taxa with morphological data ($M_{L}$) and the overall number of missing morphological characters ($M_{C}$), have a bigger effect than the amount of missing data for the fossil taxa ($M_{F}$), and when both $M_{L}$ and/or $M_{C}$ reach 50\% of missing data, the matrix does not contain enough phylogenetic information for the fossil taxa to be placed with confidence in the tree (Figure 6).

@@

\item{\textcolor{blue}{\textbf{line 557-558:} '…than the "best" maximum likelihood tree'}}

We modified the sentence as suggested : "than the ``best'' Maximum Likelihood tree" (line @@@).

\item{\textcolor{blue}{\textbf{line 549-565:} again, I am left wondering how much the differences in performance of max. likelihood and Bayesian inference techniques are driven by the use of the 'true' tree as a starting tree in the Bayesian MCMC search, while the likelihood analysis had no such guide tree to begin with. The authors at least need to make this point clear, if not re-run some subset of analyses without introducing a starting tree to test the effects of using the 'true' tree as a starting tree or not. I understand the computational limitations, but the practical limitation that empiricists will never have the 'true' tree to start with may (or may not) have a large impact on our interpretation and the generality of these findings. }}

We do understand the reviewers concern and added an analysis in the Appendix A (@@@) where we tested the effect of using the ``true'' tree as a starting tree. We found that there is no significant effect between using the ``true'' tree or a random tree (MrBayes default) as a starting tree on topological recovery. We refer to this supplementary analysis in the main text as follows:

"Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Figure 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference (see Methods) had no significant effect on topological recovery (see Appendix A - @@@ for details on the analysis of the effect of the starting tree in Bayesian inference)."

@@

\item{\textcolor{blue}{\textbf{line 599:} remove extra close parenthesis inside parenthetical statement}}

We fixed this typo.

\item{\textcolor{blue}{\textbf{line 595-606:} this paragraph seems unsatisfactory advice for practitioners.
First, advising that the tree topology be fixed using the Bayesian consensus tree before conducting analyses such as dating strongly conditions the results on only a single tree and as the authors point out in the parenthetical statement, the dating information could improve accuracy. Indeed one of the many advantages of tip-dating is the joint inference of tree topology and divergence times.
The following statements that the posterior distribution should not be discarded does not leave the reader with a clear path that they should follow - if only the fixed tree topology is used to estimate divergence times, how can the substitution rate-scaled posterior distribution of highly probable trees be used for most comparative analyses which require time-scaled branch lengths?
This paragraph requires further justification and this study does have important bearing on a subject that gets little attention. }}

We revised the statements in the whole paragraph as follows:
"Another practical implication of our results regards the tree inference methods.
Because the Bayesian consensus trees consistently recovered topologies closer to the ``best'' tree topology than the Maximum Likelihood trees, we advice using the Bayesian consensus trees as a topological constraint for tree inferences using the Total-Evidence method such as tip-dating (e.g. Ronquist et al., 2012a; Wood et al., 2013; Matzke, 2014; although it is possible that including dating information during tree inference could also improve the accuracy of the Bayesian posterior tree distribution).
Additionally, using the Bayesian consensus tree rather than the Maximum Likelihood can reduce the amount of ``false positive'' topologies.
As shown in figure 5 and discussed in the section above (Effects of tree inference methods), the Bayesian consensus tree is more likely to not resolve nodes weakly supported due to missing data than the Maximum Likelihood tree that is more likely to wrongly resolve such nodes (i.e. creating a ``false positive'' node).
Note, however, that we do not suggest discarding the Bayesian posterior tree distribution once one chose to use the Bayesian consensus tree topology even though they performed poorly in recovering the ``best'' tree topology in our simulations (this can probably be imputed to the difficulties comparing distributions of trees; see above).
This can be particularly important because these trees will be invaluable for phylogenetic comparative analyses. For example a sub-sample of posterior trees distributions can be used to asses macroecological questions while better taking into account topological uncertainty (e.g. Fritz et al. 2009 and Jetz et al. 2012 trees used in Healy et al. 2014)."
%To do (rewrite)

@@

\item{\textcolor{blue}{\textbf{line 613-615:} rephrase - sentence is choppy with all the commas. }}

We modified the sentence as follow: "Because it can be difficult to do for fossil taxa, we recommend coding as many morphological characters as possible for living taxa." (line @@@),

\item{\textcolor{blue}{\textbf{line 615-618:} It seems more important to me that the authors emphasize what implications it may have for previous and future studies that Bayesian analyses outperform max. likelihood, other than to re-hash that the authors feel the Bayesian consensus should be used to fix topologies for other downstream analyses.
For example, can the authors cite some combined analyses that used only likelihood analyses and came up with controversial results?
Or what about other studies that have done combined analyses with different optimality criteria and found different topologies?
What about implications for studies that have used molecular scaffold techniques to fix the topology of living taxa based on analyses of only DNA analyses and then fit the fossils into this fixed tree?
Also 'further' is misspelled.  }}

%Check:
%Wood
%Schrago
%Pyron
%Schrago
%Beck,
%Slater
%Ronquist
%-Wittenberg.et.al 2015 - Org Divers Evol (TEM)
%-Carrizo&Catalano 2015 - Cladistics (TEM)
%-Djernæs et al 2015 MPE (TEM)
%-Arcila et al 2015 - MPE (TEM)

@@

%To do (rewrite) - typo ok

%Original: Additionally, the topology of the Bayesian consensus trees, regardless the amount of missing data, were always closer to the “best” tree topology than the Maximum Likelihood trees. Therefore, we advise using Bayesian consensus tree topologies if topology must be fixed for furhter analyses. The results of our analyses are encouraging and show that it is possible to combine both neontological and palaeontological data in the same phylogeny despite issues of missing data. Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes.

### 2015-07-13 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
response update
Daily revision update

Here we use simulations to assess the effects of missing data on tree topologies inferred from Total Evidence matrices.
We investigate three major factors that directly affect the completeness and the size of the morphological part of the matrix: the proportion of living taxa with no morphological data, the amount of missing data in the fossil record, and the overall number of morphological characters in the matrix.
We infer phylogenies from complete matrices and from matrices with various amounts of missing data, and then compare missing data topologies to the ``best" tree topology inferred using the complete matrix.

We find that the number of living taxa with morphological characters and the overall number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the ``best" tree topology.
Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework.
Additionally, we find that Bayesian methods consistently outperform other tree inference methods.
We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses.

@@

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
These approaches differ mainly in how they treat the fossil taxa and it's data. One can use the fossil as tips or as nodes in the phylogeny and use it's age only, it's morphology only or it's age and morphology jointly.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, only fossils' occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best" approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

@@

In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by thoroughly eliminating living or fossil data or all data overall.
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa.
We focus on the effect of missing data on our ability to recover a ``best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best" tree topology more so than minimizing the amount of missing data in the fossil record.
Additionally, we find that the ability of Total Evidence methods to recover the ``best" tree topology is increased when using Bayesian methods.

@@

\section{Materials and Methods}
To explore how missing data in the morphological partition of Total Evidence matrices influences tree topology, we used the following protocol (Figure 1):
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true" tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data" matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete" matrix and from the 125 ``missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best" tree to the ``missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

%\begin{figure}[h]
%\caption{{\bf Protocol outline.}
%(1) We randomly generated a birth-death tree (the ``true" tree) and used it to simulate a matrix with no missing data (the ``complete" matrix).
%(2) We removed data from the morphological part of the ``complete" matrix resulting in 125 ``missing-data" matrices.
%(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
%(4) We compared the ``missing-data" trees to the ``best" tree.
%We repeated these four steps 50 times.}
%\label{Fig_Outline}
%\end{figure}

@@

\subsection{Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}.
We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution (bounded between 0 and 1) but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}.
Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa.
For example, fossil taxa make up 88\% \citep{beckancient2014}, 58\% \citep{schragocombining2013}, 48\% \citep{pyrondivergence2011}, 31\% \citep{ronquista2012} and 31\% \citep{slaterphylogenetic2013} of taxa in various studies.
To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa.
The fossil taxa were considered as unique tips at the end of extinct lineages.
We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true" tree.
The molecular matrix was simulated from the ``true" tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5 \citep{yangamong-site1996}.
In practice, such a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal \citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below).
%We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. %Justify
%We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years).
We selected this model and parameters settings to be a compromise between realism from empirical datasets \citep[e.g.][]{kellymolecular2014} and models with few parameters \citep[cf.][]{springermacroevolutionary2012} to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).
All the molecular information for fossil taxa was replaced by missing data ("?").

We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa.
We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).
We then ran an independent discrete character simulation for each character using the ``true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

Finally, we combined the morphological and molecular matrices obtained from the ``true" tree. Hereafter we call this the ``complete" matrix, i.e. the matrix with no missing data except for the molecular data of the fossil taxa.
%---------------------------------------------
%       2-Removing data
%---------------------------------------------

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete" matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
\end{enumerate}

@@

We created matrices using all parameter combinations resulting in 125 ($5^3$) matrices.
Note that one of these combinations has no missing data so is equivalent to the ``complete" matrix, thus we have one effectively complete matrix in our 125 ``missing-data" matrices.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.

@@

\subsection{Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best" tree inferred from the ``complete" matrix and the ``missing-data" trees inferred from the 125 matrices with various amounts of missing data.
The ``true" tree was used to generate the ``complete" matrix and reflects the ``true" evolutionary history in our simulations.
The ``best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods.
In real world situations, the ``true" tree is never available to us because we cannot know the true evolutionary history of a clade \citep[except in very rare circumstances, e.g.][]{rozen2005}.
Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best" tree, rather than the ``true" tree, as the ``best" tree is generally what we have to work with.

\subsubsection{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealp{tavare1986}; default GTRGAMMA in RAxML v. 8.0.20; \citealp{Stamatakis21012014}).
For the morphological data, we used the M\textit{kv} model \citep{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{kv} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20; \citealp{Stamatakis21012014}).
We used RAxML because it automatically corrects for acquisition bias \citep{lewisa2001}. It is also heavily used in the literature for Maximum Likelihood tree inference \citep[e.g.][]{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014} and is one of the fastest methods available \citep{Stamatakis01102008}.

>>>>

Here we use simulations to assess the effects of missing data on tree topologies inferred from Total Evidence matrices.
We investigate three major factors that directly affect the completeness and the size of the morphological part of the matrix: the proportion of living taxa with no morphological data, the amount of missing data in the fossil record, and the overall number of morphological characters in the matrix.
We infer phylogenies from complete matrices and from matrices with various amounts of missing data, and then compare missing data topologies to the ``best'' tree topology inferred using the complete matrix.

We find that the number of living taxa with morphological characters and the overall number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the ``best'' tree topology.
Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework.
Additionally, we find that Bayesian methods consistently outperform other tree inference methods.
We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses.

@@

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
These approaches differ mainly in how they treat the fossil taxa and it's data. One can use the fossil as tips or as nodes in the phylogeny and use it's age only, it's morphology only or it's age and morphology jointly.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, only fossils' occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best'' approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

@@

In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by thoroughly eliminating living or fossil data or all data overall.
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa.
We focus on the effect of missing data on our ability to recover a ``best'' tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

@@

\section{Materials and Methods}
To explore how missing data in the morphological partition of Total Evidence matrices influences tree topology, we used the following protocol (Figure 1):
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true'' tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete'' matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete'' matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data'' matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete'' matrix and from the 125 ``missing-data'' matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best'' tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data'' trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best'' tree to the ``missing-data'' trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

%\begin{figure}[h]
%\caption{{\bf Protocol outline.}
%(1) We randomly generated a birth-death tree (the ``true'' tree) and used it to simulate a matrix with no missing data (the ``complete'' matrix).
%(2) We removed data from the morphological part of the ``complete'' matrix resulting in 125 ``missing-data'' matrices.
%(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
%(4) We compared the ``missing-data'' trees to the ``best'' tree.
%We repeated these four steps 50 times.}
%\label{Fig_Outline}
%\end{figure}

@@

\subsection{Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true'' tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}.
We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution (bounded between 0 and 1) but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}.
Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa.
For example, fossil taxa make up 88\% \citep{beckancient2014}, 58\% \citep{schragocombining2013}, 48\% \citep{pyrondivergence2011}, 31\% \citep{ronquista2012} and 31\% \citep{slaterphylogenetic2013} of taxa in various studies.
To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa.
The fossil taxa were considered as unique tips at the end of extinct lineages.
We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true'' tree.
The molecular matrix was simulated from the ``true'' tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5 \citep{yangamong-site1996}.
In practice, such a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal \citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below).
%We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. %Justify
%We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years).
We selected this model and parameters settings to be a compromise between realism from empirical datasets \citep[e.g.][]{kellymolecular2014} and models with few parameters \citep[cf.][]{springermacroevolutionary2012} to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).
All the molecular information for fossil taxa was replaced by missing data ("?").

We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa.
We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

Finally, we combined the morphological and molecular matrices obtained from the ``true'' tree. Hereafter we call this the ``complete'' matrix, i.e. the matrix with no missing data except for the molecular data of the fossil taxa.
%---------------------------------------------
%       2-Removing data
%---------------------------------------------

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete'' matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
\end{enumerate}

@@

We created matrices using all parameter combinations resulting in 125 ($5^3$) matrices.
Note that one of these combinations has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.

@@

\subsection{Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best'' tree inferred from the ``complete'' matrix and the ``missing-data'' trees inferred from the 125 matrices with various amounts of missing data.
The ``true'' tree was used to generate the ``complete'' matrix and reflects the ``true'' evolutionary history in our simulations.
The ``best'' tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods.
In real world situations, the ``true'' tree is never available to us because we cannot know the true evolutionary history of a clade \citep[except in very rare circumstances, e.g.][]{rozen2005}.
Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best'' tree, rather than the ``true'' tree, as the ``best'' tree is generally what we have to work with.

\subsubsection{Maximum Likelihood}
The ``best'' tree and the ``missing-data'' trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealp{tavare1986}; default GTRGAMMA in RAxML v. 8.0.20; \citealp{Stamatakis21012014}).
For the morphological data, we used the M\textit{kv} model \citep{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{kv} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20; \citealp{Stamatakis21012014}).
We used RAxML because it automatically corrects for acquisition bias \citep{lewisa2001}. It is also heavily used in the literature for Maximum Likelihood tree inference \citep[e.g.][]{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014} and is one of the fastest methods available \citep{Stamatakis01102008}.
